Species motif participation provides unique information about species risk of extinction

### 2023-06-06 manuscript/JAE_reply.tex
syncing quotes

        I feel that this study includes a lot of work and analysis that is meaningful. For example, at line 195-196, the authors state “This suggests that network motif profiles may be more useful than global network properties when predicting mean extinction risk of species in a network.”. To me, this sentence highlights how their results fit into an up-and-coming area of ecological research: monitoring the likelihood of consumer persistence in efficient ways. Nonetheless, I found the structure of the writing in the introduction and discussion difficult to follow and some of the key points from the results were not easy to interpret because the authors have so many different conceptual ideas incorporated throughout the article.

@@

            \item Summary: I found that the ideas explored in the paper were somewhat unclear from the summary. Is the overall goal of the study to understand participation in a motif (what I understood from the abstract and title) or is it to understand how the presence of particular sub-webs (i.e., motifs) influences persistence? It seems like it is both from the results presented, but the authors highlight the former extensively and leave the latter as more of an after-thought. This may be addressed by simply taking out words like “additionally” from the start of the last few summary points.

@@

                \textbf{R:} An initial extinction of a species (i.e. the primary extinction) can, since species are connected to each other via  consumer--resource interactions, cause other species to go extinct (i.e. secondary extinctions). We have now clarified this on line XX:

                \begin{quotation}
                The loss of a species in a food web can, due to the mutual dependencies among species, trigger a cascade of additional extinctions, so called secondary extinctions. A species' risk of going secondarily extinct is known to depend on its position within a food web~\citep{Santos2021,curtsdotter2011robustness, dunne2009cascading, Eklof2006}.
                \end{quotation}

@@

             \textbf{R:}
             We have now clarified how we in this manuscript define a disturbance. This is done on line XX.

             \begin{quotation}
             Here, we focus on persistence after the loss of plants or other basal resources, i.e., the effects of bottom-up disturbances.
            \end{quotation}

@@

                \textbf{R:} We agree with the reviewer that this was not so clearly expressed. We have now re-formulated this section and clarified with an example, beginning at line XX.

                \begin{quotation}
                However, unlike in-degree and trophic level, which focus only on direct prey of the focal species or on the lengths of paths between the focal species and basal resources, motif participation also incorporates information on the focal species' interaction partners. Two species might both have three prey and a trophic level of 3, but they might have quite different motif participation if one species's prey are strict specialists and the others' prey are generalists. Importantly, since specialists are generally more vulnerable to extinction than generalists, the predator of specialists will be more likely to lose prey and go extinct itself than a predator of generalists.
                \end{quotation}

            \item Line 48: This paragraph contains important information to understanding the concept of motifs; however, I found that the description of motif participation is vague and difficult to interpret in the context of this study. For example, the “second-step indirect interactions the species are affected by, thereby capturing important structures that are missed by global and local properties” does not provide any details that would help the reader understand why motif participation is a relevant metric. Specifically, what important structures are the authors referring to here, and how are they captured by the second-step indirect interactions? Further, is there a better way to describe what a second-step indirect interaction is? I am unsure what the authors mean by this.

                \textbf{R:} To avoid introducing extra terminology, we have replaced all instances of `global' and `local' with `large-scale' and `small-scale' respectively. We hope it is now easier to understand motifs as a meso-scale descriptor of network structure. See line XX, quoted below, for a short description of motifs and definition of meso-scale structure. We have also, as described above, explained how motif participation captures information about a focal species' interaction partners, and thereby the potential for indirect interactions. We do not refer to second-step indirect interactions in order to avoid confusion with that term.

@@

                \begin{quotation}
                We generated simulated networks using the niche model, which has been shown to recreate the structure of empirical networks well~\citep{Williams2000,Stouffer2007}.
                To capture a range of network architectures similar to those in well-studied empirical networks~\citep{Dunne2002,Dunne2002e}, we simulated networks with sizes (S) ranging from 50 to 100 species (in steps of 10) and connectances (C) ranging from 0.02 to 0.18 (in steps of 0.04). 
                All networks were generated using the function ``nichemodel'' within the Julia~\citep{Bezanson2017julia} package \emph{BioEnergeticFoodWebs}~\citep{bioenergfw,Delmas2017}.   
                For full details and a comparison to empirical networks, see \emph{Appendix S2}.
                \end{quotation}

@@

                We have revised this line (quoted below) and hope it is now clearer that a species' motif participation role refers to that same species' appearance in different motifs, and that this applies to any species one selects.
                \begin{quotation}
                A species' ``motif participation role'' is the frequency with which that species appears in each of the motifs present in a network~\citep{Stouffer2012}.
                \end{quotation}

@@

                \textbf{R:} The size of a motif participation vector is always related to the number of motifs considered, not the number of species (there is one vector per species). We only considered three-species motifs in this study, as stated in lines XX, XX, and XX. Since there are only four \emph{unique} three-species motifs in these acyclic networks (a general feature of all acyclic networks), motif participation vectors are four-dimensional for all species. This will be true for any motif participation roles calculated using three-species motifs and acyclic networks, unless one included placeholders of 0 for the 9 three-species motifs containing cycles (see~\citep{Stouffer2007} for a full list of these motifs). In that case all motif participation vectors would be 13-dimensional, but still only 4 dimensions would be informative.

                We have rephrased this line (quoted below) to remove the phrase `in our case' and make it clearer that four-dimensional motif participation vectors are a general feature of acyclic networks.
                \begin{quotation}
                    As only four unique three-species motifs can appear in acyclic networks, these are four-dimensional vectors. 
                \end{quotation}

@@

                \begin{quotation}

                    There are limitations to the approach presented here.
                    Most importantly, the Bayesian network framework operates on a strict bottom-up principle and therefore cannot include top-down effects in the extinction risk calculations~\citep{Eklof2013}. 
                    Here our focus was to analyze bottom-up affects following disturbances to the basal resources, in which case the Bayesian network approach captures a majority of the secondary extinctions obtained in a fully-fledged dynamical model with much greater efficiency~\citep{Eklof2013}.
                    If top-down effects are important for a particular empirical system, however, our simulations may not be accurate. 
                    For example, although a recent study using a dynamical model to relate motif participation to persistence after a disturbance found similar trends for omnivory, participation in the direct competition motif was not universally positive~\citep{Cirtwill2022Oikos}.
                    In particular, species which frequently appeared in the prey position of the direct competition motif tended to have shorter persistence times after disturbance in that study - possibly due to top-down effects from the pair of predators involved.
                    However, as these two studies applied different types of disturbances, further study would be required to disentangle the effects of top-down and bottom-up effects on species with different motif participation, and how each relates to the type and strength of disturbance.

>>>>

        I feel that this study includes a lot of work and analysis that is meaningful. For example, at line 195-196, the authors state ``This suggests that network motif profiles may be more useful than global network properties when predicting mean extinction risk of species in a network.''. To me, this sentence highlights how their results fit into an up-and-coming area of ecological research: monitoring the likelihood of consumer persistence in efficient ways. Nonetheless, I found the structure of the writing in the introduction and discussion difficult to follow and some of the key points from the results were not easy to interpret because the authors have so many different conceptual ideas incorporated throughout the article.

@@

            \item Summary: I found that the ideas explored in the paper were somewhat unclear from the summary. Is the overall goal of the study to understand participation in a motif (what I understood from the abstract and title) or is it to understand how the presence of particular sub-webs (i.e., motifs) influences persistence? It seems like it is both from the results presented, but the authors highlight the former extensively and leave the latter as more of an after-thought. This may be addressed by simply taking out words like ``additionally'' from the start of the last few summary points.

@@

                \textbf{R:} An initial extinction of a species (i.e. the primary extinction) can, since species are connected to each other via  consumer--resource interactions, cause other species to go extinct (i.e. secondary extinctions). We have now clarified this on lines 2-5:

                \begin{quotation}
                A single extinction can, due to the mutual dependencies among species in a food web,  trigger a cascade of additional (secondary) extinctions. 
                A species' risk of secondary extinction depends on its position within a food web~\citep{Santos2021,dunne2009cascading, Eklof2006}.
                \end{quotation}

@@

             \textbf{R:}
             We have now clarified how we in this manuscript define a disturbance. This is done on line 71 (quoted below) and in the summary.

             \begin{quotation}
             Here, we analyse whether a consumer's risk of secondary extinction after removal of basal resources depends on the consumer's motif participation and how this relationship varies with the severity of disturbance.
            \end{quotation}

@@

                \textbf{R:} We agree with the reviewer that this was not so clearly expressed. We have now re-formulated this section and clarified with an example, beginning at line 21.

                \begin{quotation}
                    Importantly, motif participation incorporates information on the positions of focal species' interaction partners.
                    Two species might have the same trophic level and in-degree but quite different motif participation if one species's prey are strict specialists and the others' prey are generalists (Fig. 1).
                    Since specialists are usually more vulnerable to extinction than generalists, the predator of specialists will be more likely to lose prey and go extinct itself than will a predator of generalists.
                    This extra information on the focal species' interaction partners may thus provide extra insight when predicting extinction risk.
                \end{quotation}

            \item Line 48: This paragraph contains important information to understanding the concept of motifs; however, I found that the description of motif participation is vague and difficult to interpret in the context of this study. For example, the ``second-step indirect interactions the species are affected by, thereby capturing important structures that are missed by global and local properties'' does not provide any details that would help the reader understand why motif participation is a relevant metric. Specifically, what important structures are the authors referring to here, and how are they captured by the second-step indirect interactions? Further, is there a better way to describe what a second-step indirect interaction is? I am unsure what the authors mean by this.

                \textbf{R:} To avoid introducing extra terminology, we have replaced all instances of `global' and `local' with `large-scale' and `small-scale' respectively. We hope it is now easier to understand motifs as a meso-scale descriptor of network structure. See line 13, quoted below, for a short description of motifs and definition of meso-scale structure. We have also, as described above, explained how motif participation captures information about a focal species' interaction partners, and thereby the potential for indirect interactions. We do not refer to second-step indirect interactions in order to avoid confusion with that term.

@@

                \begin{quotation}
                Simulated networks were generated using the niche model, which recreates the structure of empirical networks well~\citep{Williams2000,Stouffer2007}.
                To capture a range of network architectures similar to those in well-studied empirical networks~\citep{Dunne2002,Dunne2002e}, networks with sizes (S) ranging from 50 to 100 species (in steps of 10) and connectances (C) ranging from 0.02 to 0.18 (in steps of 0.04) were simulated. 
                Biologically unlikely networks (i.e., those with extremely long paths between any consumer and basal resources~\citep{Borrelli2014} were removed and replaced with new simulated networks until 100 suitable networks in each combination of size and connectance were obtained.
                All networks were generated using the function ``nichemodel'' within the Julia~\citep{Bezanson2017julia} package \emph{BioEnergeticFoodWebs}~\citep{bioenergfw,Delmas2017}.     
                For full details and a comparison to empirical networks, see \emph{Appendix S1}.
                \end{quotation}

@@

                We have revised this line (quoted below) and hope it is now clearer that a species' motif participation role refers to that same species' appearance in different motifs, and that this applies to any species one selects.
                \begin{quotation}
                A species' motif participation (Fig. 1) is the vector of frequencies with which it appears in each type of motif present in a network~\citep{Stouffer2012}.
                \end{quotation}

@@

                \textbf{R:} The size of a motif participation vector is always related to the number of motifs considered, not the number of species (there is one vector per species). We only considered three-species motifs in this study, as stated in the summary and lines 28 and 95. Since there are only four \emph{unique} three-species motifs in these acyclic networks (a general feature of all acyclic networks), motif participation vectors are four-dimensional for all species. This will be true for any motif participation roles calculated using three-species motifs and acyclic networks, unless one included placeholders of 0 for the 9 three-species motifs containing cycles (see~\citep{Stouffer2007} for a full list of these motifs). In that case all motif participation vectors would be 13-dimensional, but still only 4 dimensions would be informative.

                We have rephrased line 95 (quoted below) to remove the phrase `in our case' and make it clearer that four-dimensional motif participation vectors are a general feature of acyclic networks.
                \begin{quotation}
                    As only four unique three-species motifs can appear in acyclic networks, these are four-dimensional vectors. 
                \end{quotation}

@@

                \begin{quotation}
    There are limitations to the approach presented here.
    Most importantly, the Bayesian network framework operates on a strict bottom-up principle and therefore cannot include top-down effects in the extinction risk calculations~\citep{Eklof2013}. 
    Here our focus was to analyze bottom-up affects following disturbances to the basal resources, in which case the Bayesian network approach captures a majority of the secondary extinctions obtained in a fully-fledged dynamical model with much greater efficiency~\citep{Eklof2013}.
    If top-down effects are important for a particular empirical system, however, our simulations may not be accurate. 
    For example, although a recent study using a dynamical model to relate motif participation to persistence after a disturbance found similar trends for omnivory, participation in the direct competition motif was not universally positive~\citep{Cirtwill2022Oikos}.
    In particular, species which frequently appeared in the prey position of the direct competition motif tended to have shorter persistence times after disturbance in that study - possibly due to top-down effects from the pair of predators involved.
    However, as these two studies applied different types of disturbances, further study would be required to disentangle the effects of top-down and bottom-up effects on species with different motif participation, and how each relates to the type and strength of disturbance.
    Another potentially useful addition to future work would be to include species traits such as body size, which are related to both motif participation~\citep{cirtwill2018feeding} and extinction risk~\citep{Brose2017,Cardillo2005,Purvis2000} in empirical webs.
